\section*{Suggested Solutions}

\subsection*{Section 1 --- Multiple Choice}

\begin{center}
\begin{tabular}{cccccccccc}
1:b & 2:a & 3:b & 4:b & 5:b & 6:b & 7:b & 8:c & 9:c & 10:a \\
11:a & 12:b & 13:a & 14:b & 15:b & 16:a & 17:b & 18:a & 19:a & 20:a
\end{tabular}
\end{center}

\vspace{0.5cm}

\subsection*{Section 2 --- Short Answers}

\begin{enumerate}[leftmargin=0.9cm]

\item 
\[
L_{\mathrm{FSPL}} = 32.44 + 20\log_{10}(f_{\mathrm{MHz}}) + 20\log_{10}(d_{\mathrm{km}})
\]
Frequency $\uparrow$ $\Rightarrow$ loss $\uparrow$, distance $\uparrow$ $\Rightarrow$ loss $\uparrow$.

\item 
Licensed: exclusive use (e.g., NB-IoT).  
Unlicensed: shared (e.g., LoRa/WiFi).  

a) More users $\Rightarrow$ more interference.  
b) Example: CSMA / listen-before-talk.

\item 
\[
\mathrm{SINR} = \frac{S}{N+I}
\]
Includes interference, more realistic than SNR.

\item 
Increasing bandwidth increases noise power $\Rightarrow$ SNR decreases $\Rightarrow$ diminishing capacity gain.

\item 
Example: eavesdropping or spoofing.  
Solution: encryption or authentication.

\item 
Beamforming focuses energy $\Rightarrow$ higher SNR, less interference.  
Limitation: requires channel knowledge, sensitive to mobility.

\end{enumerate}

\vspace{0.5cm}

\subsection*{Section 3 --- Long Questions}

\subsubsection*{Question 1}

a)
\[
L = 32.44 + 20\log_{10}(800) \approx 90.5\ \mathrm{dB}
\]

b)
\[
P_r = 15 + 2 + 2 - 90.5 = -71.5\ \mathrm{dBm}
\]

c)
\[
P_{\min} = -120 + 3 = -117\ \mathrm{dBm}
\]

d) Feasible ($-71.5 > -117$)

e)
\[
L_{3\,\mathrm{GHz}} \approx 102\ \mathrm{dB}, \quad P_r \approx -83\ \mathrm{dBm}
\]

f) Higher frequency $\Rightarrow$ higher path loss.

\vspace{0.3cm}

\subsubsection*{Question 2}

a)
\[
L_{\text{total}} = 75 + 20 + 3 + 2 = 100\ \mathrm{dB}
\]

b)
\[
P_r = 10 - 100 = -90\ \mathrm{dBm}
\]

c) Not feasible ($-90 < -85$)

d) Loss depends on material properties and frequency.

\vspace{0.3cm}

\subsubsection*{Question 3}

a) $N = 7$

b) $(i,j) = (2,1)$

c)
\begin{itemize}
\item Bandwidth per cell increases as $N$ decreases
\item Interference increases as $N$ decreases
\end{itemize}

\vspace{0.3cm}

\subsubsection*{Question 4}

a)
\[
N = 10^{-12}\ \mathrm{mW}, \quad I \approx 1.58\times 10^{-11}\ \mathrm{mW}
\]

b)
\[
N+I \approx 1.68\times10^{-11} \Rightarrow -107.7\ \mathrm{dBm}
\]

c)
\[
\mathrm{SINR} = -95 - (-107.7) = 12.7\ \mathrm{dB}
\]

d) Feasible ($12.7 > 5$)

e) Many users $\Rightarrow$ interference dominates.

\vspace{0.3cm}

\subsubsection*{Question 5}

a)
\begin{itemize}
\item $1 \rightarrow 2 \rightarrow G$
\item $1 \rightarrow 2 \rightarrow 3 \rightarrow G$
\item $1 \rightarrow 3 \rightarrow G$
\end{itemize}

b)
\[
5.1,\quad 5.5,\quad 6.8
\]

c) Minimum: $1 \rightarrow 2 \rightarrow G$

d) Shorter hops reduce path loss.

e) Advantage: lower power. Disadvantage: delay/complexity.

\vspace{0.3cm}

\subsubsection*{Question 6}

a)
\[
C = 200\,\mathrm{kHz} \cdot \log_2(11) \approx 200 \cdot 3.46 = 692\ \mathrm{kbps}
\]

b)
MCS $\approx$ QPSK, $R=1/2$, throughput $\approx 1$ bit/s/Hz

c)

Threshold $\approx 8$ dB

\[
\Delta = 12 - 8 = 4\ \mathrm{dB}
\]

\[
\text{Power factor} = 10^{4/10} \approx 2.51
\]

\[
P_{\min} \approx \frac{100}{2.51} \approx 40\ \mathrm{W}
\]

Reduction $\approx 60\ \mathrm{W}$

\vspace{0.3cm}

\subsubsection*{Question 7}

a) Place BS at elevated position for LOS.

b) Two-site/relay improves coverage in shadowed regions.

c)
\begin{itemize}
\item Two-site: better coverage, lower tower
\item Single-site: cheaper but more shadowing
\end{itemize}

d) Two-site better: load split, less interference/medium access contention.

e) Path loss, shadowing, diffraction.
