\documentclass[11pt,a4paper]{article}

\usepackage[margin=2.2cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb}
\usepackage{enumitem}
\usepackage{array}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{lastpage}
\usepackage{fancyhdr}
\usepackage{placeins}
\usepackage{setspace}
\usepackage{parskip}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{tikz}

\begin{document}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
\fancyfoot[R]{\thepage/\pageref{LastPage}}
\fancypagestyle{plain}{
    \fancyhf{}
    \fancyfoot[R]{\thepage/\pageref{LastPage}}
    \renewcommand{\headrulewidth}{0pt}
    \renewcommand{\footrulewidth}{0pt}
}

\title{5731.25 Wireless Communication --- Trial Exam 4}
\date{26. March 2026}
\maketitle

\section*{Instructions to Students}
\begin{itemize}
    \item Ensure clarity and precision in your answers.
    \item Attempt all sections. Each of the three sections contributes one-third (1/3) of the total mark.
    \item All answers must be written in \textbf{English}.
    \item Write your answers in the provided answer sheets using a pen.
    \item Each page must contain your name and student number.
    \item Number all submitted pages as \textit{Page X/Y}.
    \item For the multiple-choice section, indicate your answers using the format: Q1 A, Q2 C, ...
    \item In the appendix of this document, you will find a logarithm table for reference.
\end{itemize}

\newpage
\input{multiple}
\newpage

\section*{Section 2 --- Short Questions (6 questions)}
Answer briefly but clearly.

\begin{enumerate}[leftmargin=0.9cm]
\item Why does the required bandwidth $B$ increase when the symbol rate $R_s$ increases? (Explain the intuition.)

\item A symbol is transmitted over a finite interval $T_s$. State the relationship between $R_s$ and $T_s$.

\item If each symbol carries $\log_2(M)$ bits, compute the number of bits per symbol when $M=8$.

\item Define spectral efficiency $\eta = \frac{R_b}{B}$ and state what it measures.

\item Define duplexing.

\item Resource allocation: list one reason why interference management is difficult when multiple users access the network simultaneously.
\end{enumerate}

\newpage
\section*{Section 3 --- Long / Applied Questions}
\textbf{Show all calculations and reasoning.}

\subsection*{Question 1 --- Link Budget and Feasibility}
A transmitter operates at $f = 850$ MHz with:
\begin{itemize}[leftmargin=1.2cm]
    \item transmit power $P_t = 18$ dBm,
    \item transmit antenna gain $G_t = 2$ dBi,
    \item receive antenna gain $G_r = 0$ dBi,
    \item distance $d = 2.5$ km.
\end{itemize}

Assume:
\[
N = -110\ \mathrm{dBm}, \qquad \gamma_{\min} = 7\ \mathrm{dB}
\]

\begin{enumerate}[label=\alph*)]
    \item Compute the free-space path loss at 850 MHz.
    \item Compute the received power $P_r$.
    \item Compute the minimum required received power for feasibility.
    \item Is the link feasible?
\end{enumerate}

\subsection*{Question 2 --- Indoor Path Loss}
An indoor link has:
\begin{itemize}[leftmargin=1.2cm]
    \item distance-based path loss of $72$ dB,
    \item two concrete walls causing $9$ dB loss each,
    \item one wooden door causing $2$ dB loss,
    \item one glass window causing $4$ dB loss.
    \end{itemize}

Transmitter and receiver antenna gains are $0$ dBi, and $P_t = 12$ dBm.

\begin{enumerate}[label=\alph*)]
    \item Compute total path loss.
    \item Compute received power.
    \item If receiver sensitivity is $-83$ dBm, does the link work?
    \item Suggest two simple ways to improve this link without changing bandwidth.
\end{enumerate}

\subsection*{Question 3 --- Frequency Reuse}
Assume a cellular layout with reuse factor $N=5$.

\begin{enumerate}[label=\alph*)]
    \item What fraction of total spectrum is available in each cell?
    \item Compare this with a system using $N=8$.
    \item Explain how changing $N$ impacts:
    \begin{enumerate}[label=\roman*)]
        \item co-channel interference,
        \item capacity per cell.
    \end{enumerate}
\end{enumerate}

\subsection*{Question 4 --- Multi-hop Routing}
Only the following links are feasible:
\[
1 \rightarrow 2,\quad 1 \rightarrow 3,\quad 2 \rightarrow 3,\quad 2 \rightarrow 4,\quad 3 \rightarrow 4,\quad 1 \rightarrow 4
\]
Transmission energy costs are:
\[
E_{12}=1.8,\quad E_{13}=3.2,\quad E_{23}=1.1,\quad E_{24}=2.2,\quad E_{34}=2.0,\quad E_{14}=6.4
\]
where node 1 is source and node 4 is gateway.

\begin{enumerate}[label=\alph*)]
    \item List all feasible routes from node 1 to node 4.
    \item Compute total energy for each route.
    \item Identify the minimum-energy route.
    \item Give one advantage and one disadvantage of multi-hop communication.
\end{enumerate}

\subsection*{Question 5 --- Modulation and Throughput}
A wireless channel has bandwidth $B = 120$ kHz.

Figure~\ref{fig:amc_snr_2} shows throughput (bits/s/Hz) versus SNR for several candidate modulation/coding schemes.
\input{../trial_exam_2/figure}
\FloatBarrier

\begin{enumerate}[label=\alph*)]
    \item For $\mathrm{SNR}=6$ dB, compute Shannon capacity in bit/s.
    \item From Figure~\ref{fig:amc_snr_2}, choose the highest usable scheme at 6 dB and compute data rate in bit/s.
    \item At 7 dB the system uses scheme-3. How much SNR drop (in dB) is possible before switching down to scheme-2?
\end{enumerate}

\section*{Question 6 --- Coverage Planning}
The terrain sketch on the last page shows two towns separated by a hill.

\begin{enumerate}[label=\alph*)]
\item Propose a single base-station location and explain your choice.
\item Propose a two-site solution (base station + relay) and explain the benefit.
\item Compare the two solutions in terms of coverage, cost, and reliability.
\item Which propagation effects are most important in this scenario?
\end{enumerate}

\vspace{0.6cm}
\hrule
\vspace{0.3cm}
\textbf{End of exam}

\input{lastpages}
\clearpage
\label{LastPage}
\end{document}
