\documentclass[12pt]{article}
\usepackage{graphicx} % Required for inserting images
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{fancyhdr}
\usepackage{lastpage}  % Ensures total page count

\title{5730 - Mobile Communications and Internet of Things Spring Exam 2025}
\date{18. March 2025}

\begin{document}
\pagestyle{fancy}
\fancyhf{}  % Clear default headers and footers
\renewcommand{\headrulewidth}{0pt}  % No top horizontal line
\renewcommand{\footrulewidth}{0pt}  % No bottom horizontal line
\fancyfoot[R]{\thepage/\pageref{LastPage}}  % Right footer: X/Y format
\fancypagestyle{plain}{
    \fancyhf{}  % Clear default first-page footer
    \fancyfoot[R]{\thepage/\pageref{LastPage}}  % Apply the same footer
    \renewcommand{\headrulewidth}{0pt}  % Remove top line on first page
    \renewcommand{\footrulewidth}{0pt}  % Ensure no bottom line
}
\maketitle

\section*{Instructions to Students}

\begin{itemize}
    \item Ensure clarity and precision in your answers.
    \item Attempt all sections. Each of the three sections contributes one-third (1/3) of the total mark.
    \item All answers must be written in \textbf{English}.
    \item Write your answers in the provided answer sheets using a pen.
    \item Each page must contain your name and student number.
    \item Number all submitted pages as \textit{Page X/Y}, where \textit{X} is the page number and \textit{Y} is the total (e.g., \textit{1/9, 2/9, ..., 9/9} for nine pages). 
    \item For the multiple-choice section, indicate your answers using the format: Q1 A, Q2 C, ...
    \item In the appendix of this document, you will find a logarithm tables for reference.
\end{itemize}

\input{mcq}

\section{Short Questions}
\begin{enumerate}
    \item Briefly explain the role of eMBB in 5G communication.
    \item Define edge computing and its significance in IoT applications.
    \item What is the purpose of beamforming in massive MIMO?
    \item Mention two real-world applications of massive IoT.
    \item What is the relationship between bandwidth and data rate in a wireless communication system?
    \item Compare QPSK and 16-QAM in terms of spectral efficiency and robustness to noise. 
    \item How does massive MIMO contribute to energy efficiency in 5G networks?
    \item What is the difference between SDN and traditional network architecture?
    \item What are the advantages and disadvantages of using licensed vs. unlicensed spectrum for IoT applications? Provide one example for each.
    \item SHA-256 is widely used for cryptographic hashing and authentication in IoT security. Explain its role in authentication and discuss its limitations in IoT applications.
\end{enumerate} 
\newpage
\section{Logical Reasoning}
Consider an IoT parking sensor that must communicate with a 5G base station located 1 km away.
The sensor has a transmission power of 20 dBm, and the base station transmits at 23 dBm. The antenna gain of the sensor and base station is 0 dBi. The carrier frequency is 3 GHz. Various losses in the systems amount to 8 dB.
\begin{enumerate}
    \item The IoT device has a receive bandwidth of 180 kHz. Calculate the noise floor at room temperature (290K).
    \item Using the same IoT sensor and base station setup:
    \begin{enumerate}
        \item Compute the Free-Space Path Loss (FSPL). For ease of calculation you can assume that the speed of light $c \approx 3\cdot10^8$ and $\pi \approx 3$.
        \item If the transmitting device is aiming for an SNR of 0 dB at the receiver, calculate the required transmission power $P_t$.
        \item Explain what happens to the SNR and data rate if the bandwidth is increased to 1.4 MHz, provided that the transmission power remains as calculated in 2b. Does increasing bandwidth always improve the data rate?
    \end{enumerate}
    \item Assume the system supporting the IoT network uses 16-QAM modulation.
    \begin{enumerate}
        \item How many bits per symbol does this modulation scheme support?
        \item Would increasing the modulation order to 64-QAM improve or worsen the communication for the IoT sensor? Explain.
    \end{enumerate}
\item A 5G system has a received power that results in an SNR of 0 dB. However, the system requires an SNR of 10 dB for reliable communication.
    \begin{enumerate}
        \item Compute the number of beamforming antennas required to achieve the required SNR.
        \item Assuming a bandwidth of 100 MHz, calculate the theoretical channel capacity for SNR = 10 dB.
    \end{enumerate}

    \end{enumerate}
\newpage
\section{Supplemental material}
\subsection{Logarithm Properties}
\begin{itemize}
    \item \textbf{Product Rule:} 
    \[
    \log_b (xy) = \log_b (x) + \log_b (y)
    \]
    \item \textbf{Quotient Rule:} 
    \[
    \log_b \left( \frac{x}{y} \right) = \log_b (x) - \log_b (y)
    \]
    \item \textbf{Power Rule:} 
    \[
    \log_b (x^n) = n \log_b (x)
    \]
    \item \textbf{Base Change Formula:} 
    \[
    \log_b (x) = \frac{\log_c (x)}{\log_c (b)}
    \]
    \item \textbf{Log of Reciprocal:} 
    \[
    \log_b \left( \frac{1}{x} \right) = -\log_b (x)
    \]
    \item \textbf{Logarithm to Exponential Conversion:}
    \[
    x \cdot \log_b(y) = z \quad \Rightarrow \quad y = b^{\frac{z}{x}}
    \]
\end{itemize}
\subsection{Commonly Used Constants and Units}

\begin{itemize}
    \item \textbf{Speed of Light:} 
    \[
    c \approx 3.0 \times 10^8 \text{ m/s}
    \]
    \item \textbf{Boltzmann's Constant:} 
    \[
    k_B = 1.38 \times 10^{-23} \text{ J/K}
    \]
        \item \textbf{Thermal Noise Power Density:}  
    \[
    P_n (1 \text{ Hz}) = -174 \text{ dBm/Hz}
    \]
    
\end{itemize}

\subsection{Logarithm Conversion Examples}
\begin{itemize}
    \item \(\log_{2} (16) = 4\), since \(2^4 = 16\).
    \item \(\log_{10}(12) = \log_{10}(6)+\log_{10}(2)\)
    \item Converting \( 20 \) dB to a linear factor: \(10^{\frac{20}{10}} = 10^{2} = 100\)

\end{itemize}

\begin{table}[h]
    \centering
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{|l|r|r|r|r|}
        \hline
        \textbf{Value} & \(\log_{10}(\text{Value})\) & \(\log_{2}(\text{Value})\) & Scientific Notation & SI - Prefix \\
        \hline
        \(10^{-9}\)   & -9.000  & -29.897 & \(1.0 \times 10^{-9}\) & Nano \\
        \(10^{-8}\)   & -8.000  & -26.575 & \(1.0 \times 10^{-8}\) & \\
        \(10^{-7}\)   & -7.000  & -23.253 & \(1.0 \times 10^{-7}\) & \\
        \(10^{-6}\)   & -6.000  & -19.931 & \(1.0 \times 10^{-6}\) & Micro \\
        \(10^{-5}\)   & -5.000  & -16.609 & \(1.0 \times 10^{-5}\) & \\
        \(10^{-4}\)   & -4.000  & -13.287 & \(1.0 \times 10^{-4}\) & \\
        \(10^{-3}\)   & -3.000  & -9.966  & \(1.0 \times 10^{-3}\) & Milli \\
        \(0.01\)   & -2.000  & -6.644  & \(1.0 \times 10^{-2}\) & \\
        \(0.1\)   & -1.000  & -3.322  & \(1.0 \times 10^{-1}\) & \\
        \hline
        1             & 0.000   & 0.000   & \(1.0 \times 10^{0}\) & - \\
        2             & 0.301   & 1.000   & \(2.0 \times 10^{0}\) & \\
        3             & 0.477   & 1.585   & \(3.0 \times 10^{0}\) & \\
        4             & 0.602   & 2.000   & \(4.0 \times 10^{0}\) & \\
        5             & 0.699   & 2.322   & \(5.0 \times 10^{0}\) & \\
        6             & 0.778   & 2.585   & \(6.0 \times 10^{0}\) & \\
        7             & 0.845   & 2.807   & \(7.0 \times 10^{0}\) & \\
        8             & 0.903   & 3.000   & \(8.0 \times 10^{0}\) & \\
        9             & 0.954   & 3.170   & \(9.0 \times 10^{0}\) & \\
        10            & 1.000   & 3.322   & \(1.0 \times 10^{1}\) & \\
        \hline
        \(100\)      & 2.000   & 6.644   & \(1.0 \times 10^{2}\) & \\
        \(1000\)      & 3.000   & 9.966   & \(1.0 \times 10^{3}\) & Kilo \\
        \(10^4\)      & 4.000   & 13.287  & \(1.0 \times 10^{4}\) & \\
        \(10^5\)      & 5.000   & 16.609  & \(1.0 \times 10^{5}\) & \\
        \(10^6\)      & 6.000   & 19.931  & \(1.0 \times 10^{6}\) & Mega \\
        \(10^7\)      & 7.000   & 23.253  & \(1.0 \times 10^{7}\) & \\
        \(10^8\)      & 8.000   & 26.575  & \(1.0 \times 10^{8}\) & \\
        \(10^9\)      & 9.000   & 29.897  & \(1.0 \times 10^{9}\) & Giga \\
        \hline
    \end{tabular}
\end{table}

\end{document}
